---
title: "growth"
format: html
---

```{r}
#| label: loading-data
#| output: false

# load libraries
library(ggplot2)
library(RColorBrewer)
library(GGally)
library(tmap)
library(dplyr)
library(purrr)
library(tidyr)
library(ggalluvial)
library(sf)

# set up directories
climdir <- file.path('data', 'raw_data', 'NOAA')
sourcedir <- file.path('data', 'raw_data')
targetdir <- file.path('data', 'derived_data')

# load cities
nodes <- st_read(file.path(targetdir, 'nodes.gpkg'))
nodes$Town_First_Mentioned <- as.numeric(nodes$Town_First_Mentioned)
nodes$Settlement_From <- as.numeric(nodes$Settlement_From)

# load demographic data
population <- read.csv(file = file.path(sourcedir, 'Viabundus-1.3-CSV',  'Population.csv'))
# Load modern countries for background maps
w <- rnaturalearthhires::map_units10[rnaturalearthhires::map_units10$REGION_UN == 'Europe',] |>
  st_transform(3035) |>
  st_crop(st_buffer(nodes, 1.5e+05))
# combine population and urban data
spatial_pop <- merge(population,  nodes, by.x = 'Nodes_ID', by.y = 'ID', all.x = TRUE)
# reset it as spatial
spatial_pop <- st_as_sf(spatial_pop)
# Load linguistic areas
areas <- st_read(dsn  = file.path('data', 'derived_data', 'linguistic_areas.gpkg') ) |>
  st_transform(st_crs(spatial_pop))

```

## HYDE data

The folder is organised in a way that there are base, upper and lower estimates. Then, rasters are divided by year.

The innermost folder with the year estimate contains:

popc_<yr>.asc (population counts, in inhabitants/gridcell)
popd_<yr>.asc (population density, in inhabitants/km2 per gridcell)
rurc_<yr>.asc (rural population counts, in inh/gridcell)
urb_<yr>.asc  (urban population counts, in inh/gridcell)
uopp_<yr>.asc (total built-up area, such as towns, cities, etc, in km2 per grid cell)


```{r}
#| label: load-hyde


library(terra)

# get scenarios (base, low, high)
scenarios <- list.dirs(hydedir, recursive = FALSE)

hyde <- lapply(
  X = scenarios, 
  FUN = function(scen) {
    tmp <- list.files(scen, pattern = "pop\\.zip$", full.names = TRUE)
    
    tmp_yr <- lapply(
      X = tmp,
      FUN = function(yr) {
        
        files_in_zip <- unzip(yr, list = TRUE)$Name
        # print(files_in_zip)
        
        temp <- tempdir()
        tmp_asc <- unzip(zipfile = yr, exdir = temp)
        
        # print(tmp_asc)      
        tmp_est <- lapply(
          X = seq_along(files_in_zip),
          FUN = function(i) {
            rast(tmp_asc[i]) |>
              crop(v_areas) |>
              mask(v_areas)
          }
        )
        names(tmp_est) <- unlist(strsplit(files_in_zip, "_"))[seq(1, length(files_in_zip)*2, by = 2)]
        return(tmp_est)
      }
    )
    names(tmp_yr) <- unlist(strsplit(list.files(scen, pattern = "pop\\.zip$"), "_"))[seq(1, length(tmp_yr)*2, by = 2)]
    return(tmp_yr)
    # print(tmp)
  }
)
names(hyde) <- c('base', 'upper', 'low')

pop_stats <- lapply(
  X = hyde,
  FUN = function(estim) {
    
    do.call(rbind, lapply(names(estim), function(yr) {
    
      rasters <- estim[[yr]]
      
      total_pop  <- sum(values(rasters$popc), na.rm = TRUE)
      rural_pop  <- sum(values(rasters$rurc), na.rm = TRUE)
      urban_pop  <- sum(values(rasters$urb),  na.rm = TRUE)
      
      tmp_df <- data.frame(
        year = as.integer(gsub("AD", "", yr)),
        total = total_pop,
        rural = rural_pop,
        urban = urban_pop
      )
      return(tmp_df)
    }))
  
  }
)

# combine scenarios
df_base  <- pop_stats$base  %>% mutate(scenario = "base")
df_upper <- pop_stats$upper %>% mutate(scenario = "upper")
df_low   <- pop_stats$low   %>% mutate(scenario = "low")

df_all <- bind_rows(df_base, df_upper, df_low)

# reshape to long format
df_long <- df_all %>%
  pivot_longer(cols = c(total, rural, urban),
               names_to = "type", values_to = "population")

# plot
ggplot(df_long, aes(x = year, y = population, color = type, linetype = scenario)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c(total = "black", rural = "forestgreen", urban = "firebrick")) +
  scale_linetype_manual(values = c(base = "solid", upper = "dashed", low = "dotted")) +
  labs(
    x = "Year AD",
    y = "Population (inhabitants)",
    color = "Population type",
    linetype = "Scenario",
    title = "HYDE Population Over Time (Base, Upper, Low)"
  ) +
  theme_minimal(base_size = 14)


```
### Population by region 

Same plot but divided by the 4 regions

```{r}
#| label: reg-hyde

reg_hyde <- lapply(areas$region, function(region) {
  lapply(hyde, function(estim) {
    lapply(estim, function(period) {
      # period is a list of SpatRaster
      lapply(period, function(r) {
        crop(r, v_areas[v_areas$region == region]) |>
          mask(v_areas[v_areas$region == region])
      })
    })
  })
})
names(reg_hyde) <- areas$region

reg_pop_stats <- lapply(reg_hyde, function(region_list) {
  
  # region_list: list of scenarios for one region
  lapply(region_list, function(scenario_list) {
    
    # scenario_list: list of years
    do.call(rbind, lapply(names(scenario_list), function(yr) {
      rasters <- scenario_list[[yr]]
      
      # compute totals
      total_pop <- sum(values(rasters$popc), na.rm = TRUE)
      rural_pop <- sum(values(rasters$rurc), na.rm = TRUE)
      urban_pop <- sum(values(rasters$urb), na.rm = TRUE)
      
      data.frame(
        year = as.integer(gsub("AD", "", yr)),
        total = total_pop,
        rural = rural_pop,
        urban = urban_pop,
        stringsAsFactors = FALSE
      )
    }))
    
  })
})


# Flatten reg_pop_stats (base, low, upper) into one data.frame with region column
df_regions <- bind_rows(
  lapply(names(reg_pop_stats), function(region_name) {
    region_list <- reg_pop_stats[[region_name]]
    # assume region_list contains all 3 scenarios: base, low, upper
    bind_rows(
      lapply(names(region_list), function(scenario_name) {
        df <- region_list[[scenario_name]]  # data.frame for that scenario
        df$region   <- region_name
        df$scenario <- scenario_name
        df
      })
    )
  })
)


# Pivot to long format for ggplot
df_long <- df_regions %>%
  pivot_longer(cols = c(total, rural, urban),
               names_to = "type",
               values_to = "population")

# If you have 'low' and 'upper' scenarios, compute ribbon bounds
df_ribbon <- df_long %>%
  group_by(region, year, type) %>%
  summarize(
    lower = population[scenario == "low"],
    upper = population[scenario == "upper"],
    .groups = "drop"
  )

# Merge ribbon info into base scenario for plotting lines
df_plot <- df_long %>%
  filter(scenario == "base") %>%
  left_join(df_ribbon, by = c("region", "year", "type"))

# Define colors
line_colors <- c(total = "black", rural = "forestgreen", urban = "firebrick")

# Plot
ggplot(df_plot, aes(x = year, y = population, color = type, fill = type)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = line_colors) +
  scale_fill_manual(values = line_colors) +
  facet_wrap(~region, scales = "free_y") +
  labs(
    x = "Year AD",
    y = "Population (inhabitants)",
    color = "Population type",
    fill = "Population type",
    title = "Population Over Time by Region with Uncertainty"
  ) +
  theme_minimal(base_size = 14)


```

